---
title: "Module 2"
author: "Sara Padula"
date: "2025-05-23"
output: html_document
---
## Project Goals
-Import, clean, and organize chickadee fecal metabarcoding data from 2019, 2020, and 2024
-Create phyloseq objects that link DNA-based diet data to taxonomy and sample metadata
-Convert diet data to presence/absence matrices for analyses of dietary richness and overlap
-Visualize where samples were collected and examine how sample numbers break down by species and year

## Import and Clean Metabarcoding Data

Here we load all necessary input data for fecal sample-based diet analyses. These include:
-Chickadee metadata (master_data)
-Sample metadata (coldata)
-Taxonomy tabls and ESV (extact sequence variant) abundance data from fecal metabarcoding for 2019 and 2020
```{r import-metabarcoding}
#load libraries
library(dplyr)
library(reshape2)
library(ggplot2)
library(tibble)
library(phyloseq)



# Load metadata and ESV data
chickadee     <- read.csv('master_data.csv')
sampleID      <- read.csv("ColData.csv", header = T, check.names = F)
esv           <- read.csv("ESV_long.csv", header = T, check.names = FALSE)
ESV.2019      <- read.csv("ESV_Long_2019_Ind.csv", header = T, check.names = FALSE)
ESV.2020      <- read.csv("ESV_Long_2020_Ind.csv", header = T, check.names = FALSE)
col.data      <- read.csv("ColData.csv")
row.data.2019 <- read.csv("RowData_2019.csv")
row.data.2020 <- read.csv("RowData_2020.csv")

# Load 2024 data
esv_2024 <- read.csv("JVB4821-ArthCOI-read-data.csv", header = TRUE, check.names = FALSE)
taxonomy_2024 <- read.csv("JVB4821-ArthCOI-esv-data.csv", header = TRUE)

# Transpose and binarize 2024 ESV data
sample_cols_2024 <- grep("^S", names(esv_2024), value = TRUE)
esv_2024_transposed <- setNames(data.frame(t(esv_2024[, sample_cols_2024])), esv_2024$ESVId)
rownames(esv_2024_transposed) <- sample_cols_2024
esv_2024_binary <- esv_2024_transposed
esv_2024_binary[esv_2024_binary > 0] <- 1

# Subset coldata to just 2024 samples
coldata_2024 <- subset(sampleID, grepl("^S", SampleID))

#Format ESV tables: make ESV ID the row name and convert to matrix

ESV.2019 <- ESV.2019 %>% column_to_rownames("ESV") %>% as.matrix()
ESV.2020      <- ESV.2020 %>% column_to_rownames("ESV") %>% as.matrix()

#Make taxonomy tables: ensure unique row names for each ESV
row.data.2019 <- row.data.2019 %>%
  distinct(Query, .keep_all = TRUE) %>%
  column_to_rownames("Query") %>%
  as.matrix()

row.data.2020 <- row.data.2020 %>%
  mutate(Query = make.unique(as.character(Query))) %>%
  column_to_rownames("Query") %>%
  as.matrix()

#Format sample metadata for use in phyloseq

col.data      <- col.data %>% column_to_rownames("SampleID")

#Create phyloseq-compatable components
samples   <- sample_data(col.data)
OTU.2019  <- otu_table(ESV.2019, taxa_are_rows = TRUE)
TAX.2019  <- tax_table(row.data.2019)
OTU.2020  <- otu_table(ESV.2020, taxa_are_rows = TRUE)
TAX.2020  <- tax_table(row.data.2020)

#Fix sample naming issues in 2020 ESV matrix
sample_names(OTU.2020) <- sample_names(OTU.2020) %>%
  gsub("^X", "", .) %>%
  gsub("\\.", "-", .)

#Remove row labeled "Sum" that is not a real sample
OTU.2020 <- prune_samples(sample_names(OTU.2020) != "Sum", OTU.2020)


```

## Create Phyloseq Objects
Now we create phyoseq objects for each year. These combine the ESV counts, taxonomy and sample metadata into one unified structure for downstream ecological analyses.

```{r create-phyloseq}
chickadee.diet.2019 <- phyloseq(OTU.2019, TAX.2019, samples)
chickadee.diet.2020 <- phyloseq(OTU.2020, TAX.2020, samples)

```
## Binary Matrices: Presence/Absence of Diet Items
We will also convert ESV count tables into binary matrices (1 = detected, 0 = not detected) to look at diet breadth and overlap without considering how many reads were recovered.

```{r}
# Transpose ESV tables: now rows = samples, cols = taxa
esv.long.2019 <- t(ESV.2019)

# Convert to binary: any detection becomes 1
esv.long_bin.2019 <- esv.long.2019
esv.long_bin.2019[esv.long_bin.2019 > 0] <- 1

esv.long.2020 <- t(ESV.2020)
esv.long_bin.2020 <- esv.long.2020
esv.long_bin.2020[esv.long_bin.2020 > 0] <- 1


```

Fecal Sample Map

This map shows where fecal samples were collected, grouped by species. It's useful for identifying geographic patterns in sampling that might explain dietary variation.

```{r, make sample map and figure out where samples are coming from, how they break down by year}

library(ggmap)
library(dplyr)

# Register your Google Maps API key
register_google(key = "AIzaSyCWzA4n4xK412hVo9WezJFaZibVm3KpleM")

head(chickadee)

# Generate a requested bounding box with some padding
bbox <- make_bbox(lon = long, lat = lat, data = chickadee, f = 0.1)

# Download the map
chickadeemap <- get_map(location = bbox, maptype = "satellite", source = "google", color = "bw")

# Extract actual bounding box used by Google
actual_bbox <- attr(chickadeemap, "bb")

# Filter chickadee data to only include points within that bounding box
chickadee_filtered <- chickadee %>%
  filter(lat >= actual_bbox$ll.lat,
         lat <= actual_bbox$ur.lat,
         long >= actual_bbox$ll.lon,
         long <= actual_bbox$ur.lon)

# Plot the filtered points
map.samples_species <- ggmap(chickadeemap) + 
  geom_point(data = chickadee_filtered, 
             aes(x = long, y = lat, fill = species), 
             size = 5, alpha = 0.7, shape = 21) +
  scale_fill_manual(values = c("rosybrown2", "cornflowerblue")) +
  labs(x = "Longitude", y = "Latitude", fill = "Species")

# Display the map
map.samples_species



```


## Study Summary Stats

This gives us a big-picture look at how our samples break down across species and years. This context is critical before comparing diets.
```{r summary, include=TRUE}

#Preview the metadata
head(chickadee)
str(chickadee)

#Count nests per species
chickadee %>% 
  group_by(species) %>% 
  tally()

#Count nests per year
chickadee %>% 
  group_by(year) %>% 
  tally()

#Count nests per species per year
chickadee %>% 
  group_by(year, species) %>% 
  tally()


#Repeat the same summary using sample metadata
col.data %>% 
  group_by(Year, Species) %>% 
  tally()


col.data %>% 
  group_by(Species) %>% 
  tally()

```

Fecal sample counts

Before analyzing diet composition, it's helpful to know how many fecal samples we have for each group. This helps gauge statistical power and ensures we're comparing like with like.

```{r fecal sample summary}

# Total number of fecal samples collected per species in 2019
chickadee %>% 
  group_by(species) %>% 
  filter(year == "2019") %>%
  summarize(fecal_samples = sum(num_fecal, na.rm = TRUE))

# Total number in 2020
chickadee %>% 
  group_by(species) %>% 
  filter(year == "2020") %>%
  summarize(fecal_samples = sum(num_fecal, na.rm = TRUE))

# Total number in 2024
chickadee %>% 
  group_by(species) %>% 
  filter(year == "2024") %>%
  summarize(fecal_samples = sum(num_fecal, na.rm = TRUE))

```



Reflection questions:
1. How might differences in sampling location, year, or species identity influence the diet data we will eventually analyze?
Consider environmental factors (e.g., elevation, temperature, habitat) and whether sampling was evenly distributed.

2. Why is it useful to convert diet data from read counts to presence/absence format for some analyses? What kinds of questions might each format help answer?
Think about the difference between detecting a food item and quantifying how much of it was eaten.

3.How do the beak differences you observed in Module 1 help you generate hypotheses about which insects might appear more frequently in one species’ diet than the other?
Try to connect morphology (e.g., deeper vs. longer beaks) to feeding strategies.

## Load 2024 Diet Data

```{r load-2024-data}
esv_2024 <- read.csv("JVB4821-ArthCOI-read-data.csv", header = TRUE, check.names = FALSE)
taxonomy_2024 <- read.csv("JVB4821-ArthCOI-esv-data.csv", header = TRUE)
```

## Transpose and Process 2024 Data

```{r process-2024-data}
sample_cols_2024 <- grep("^S", names(esv_2024), value = TRUE)
esv_2024_transposed <- setNames(data.frame(t(esv_2024[, sample_cols_2024])), esv_2024$ESVId)
rownames(esv_2024_transposed) <- sample_cols_2024

esv_2024_binary <- esv_2024_transposed
esv_2024_binary[esv_2024_binary > 0] <- 1
```

## Merge Taxonomy with 2024 Data

```{r merge-taxonomy-2024}
taxonomy_unique_2024 <- taxonomy_2024[!duplicated(taxonomy_2024$ESVId), ]
esv_2024_binary$SampleID <- rownames(esv_2024_binary)
esv_2024_long <- reshape2::melt(esv_2024_binary, id.vars = "SampleID", variable.name = "ESVId", value.name = "Presence")
esv_2024_merged <- merge(esv_2024_long, taxonomy_unique_2024, by = "ESVId")
esv_2024_present <- subset(esv_2024_merged, Presence == 1)
```

## Summary of 2024 Diet

```{r summary-2024}
library(dplyr)

# Richness per sample
richness_2024 <- esv_2024_present %>%
  group_by(SampleID) %>%
  summarize(Richness = n_distinct(ESVId))

# Most frequent taxa
top_taxa_2024 <- esv_2024_present %>%
  group_by(Order) %>%
  summarize(Frequency = n()) %>%
  arrange(desc(Frequency)) %>%
  head(10)

print(richness_2024)
print(top_taxa_2024)
```