---
title: "Module 4: Fitness and Conservation"
author: "Sara Padula"
date: "2025-06-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project Goals

- Visualize how diets differ between mountain and black-capped chickadees, and across elevation.
- Relate diet to reproductive outcomes.
- Explore conservation implications of diet patterns.

## Load Libraries

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(vegan)
library(ggpubr)
library(readr)
```

## Load and Process Diet Data from All Years

```{r}
# Load metadata and taxonomy
library(readr)
taxonomy <- read_csv("JVB4821-ArthCOI-esv-data.csv")

col_data <- read.csv("ColData.csv")
master_data <- read.csv("master_data.csv")

# Load ESV data
esv_2019 <- read.csv("ESV_Long_2019_Ind.csv", check.names = FALSE)
esv_2020 <- read.csv("ESV_Long_2020_Ind.csv", check.names = FALSE)
esv_2024 <- read.csv("JVB4821-ArthCOI-read-data.csv", check.names = FALSE)

# Ensure proper column names
colnames(esv_2019)[1] <- "ESVId"
colnames(esv_2020)[1] <- "ESVId"
colnames(esv_2024) <- gsub("\\.1$", "", colnames(esv_2024))

# Get 2024 sample columns
sample_cols_2024 <- grep("^S\\d+", colnames(esv_2024), value = TRUE)

# Calculate richness
process_long_esv <- function(df, id_col) {
  df_long <- df %>%
    pivot_longer(cols = -!!sym(id_col), names_to = "SampleID", values_to = "count") %>%
    filter(count > 0) %>%
    group_by(SampleID) %>%
    summarise(diet_richness = n_distinct(!!sym(id_col)), .groups = "drop")
  return(df_long)
}

rich_2019 <- process_long_esv(esv_2019, "ESVId")
rich_2020 <- process_long_esv(esv_2020, "ESVId")
rich_2024 <- colSums(esv_2024[, sample_cols_2024] > 0)
r_df_2024 <- data.frame(SampleID = names(rich_2024), diet_richness = as.numeric(rich_2024))

# Combine
richness_all <- bind_rows(rich_2019, rich_2020, r_df_2024)
richness_data <- left_join(col_data, richness_all, by = "SampleID")
```

## Richness Visualizations

```{r}
richness_data_clean <- richness_data %>%
  filter(!is.na(diet_richness), !is.na(Elevation), !is.na(Species))

p1 <- ggplot(richness_data_clean, aes(x = Species, y = diet_richness, fill = Species)) +
  geom_boxplot() +
  labs(title = "Diet Richness by Species", y = "# of Unique Prey Items") +
  theme_minimal()

p2 <- ggplot(richness_data_clean, aes(x = Elevation, y = diet_richness, color = Species)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Diet Richness Across Elevation", y = "Diet Richness") +
  theme_minimal()

ggarrange(p1, p2, ncol = 2)
```
Left plot
-MOCH have a wider range of diet richness than BCCH
-MOCH tend to eat a larger number of unique prey items than BCCH on average
-Some MOCH individuals had especially high richness, while BCCH richness was more tightly clustered around lower values

Right plot
-for both species diet richness generally increases with elevation
-MOCH show a stronger upward trend meaning they tend to eat more kinds of prey as elevation increases
-BCCH are mostly found at lower elevations and show a weaker trend


## Insect Family Composition from All Years

```{r}
longify_taxa <- function(df, id_col) {
  df_long <- df %>%
    pivot_longer(cols = -!!sym(id_col), names_to = "SampleID", values_to = "count") %>%
    filter(count > 0)
  return(df_long)
}

long_2019 <- longify_taxa(esv_2019, "ESVId")
long_2020 <- longify_taxa(esv_2020, "ESVId")
long_2024 <- esv_2024 %>%
  select(ESVId, all_of(sample_cols_2024)) %>%
  pivot_longer(cols = starts_with("S"), names_to = "SampleID", values_to = "count") %>%
  filter(count > 0)

all_taxa <- bind_rows(long_2019, long_2020, long_2024) %>%
  left_join(taxonomy[, c("ESVId", "Family")], by = "ESVId") %>%
  left_join(col_data, by = "SampleID") %>%
  filter(!is.na(Family), !is.na(Species))

top_families <- all_taxa %>%
  group_by(Family) %>%
  summarise(total = sum(count), .groups = "drop") %>%
  slice_max(order_by = total, n = 20) %>%
  pull(Family)

filtered_taxa <- all_taxa %>%
  filter(Family %in% top_families) %>%
  group_by(Species, Family) %>%
  summarise(total = sum(count), .groups = "drop") %>%
  group_by(Species) %>%
  mutate(prop = total / sum(total))

ggplot(filtered_taxa, aes(x = reorder(Family, -prop), y = prop, fill = Species)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  labs(title = "Top 20 Insect Families by Proportional Abundance", y = "Proportion", x = "Family") +
  theme_minimal()
```
We are showing the top most common insect families found in the chickadee diet across all years

Key takeaways
-Noctuidae (moth family) was the most dominant food source for both BCCH and MOCH
-Pollenidae and Aphididae (flies and aphids) were also important for BCCH but less so for MOCH
-MOCH had higher proportions of Tortricidae (leafroller moths) and Philodromidae (running crab spiders)
-some insect families (linyphiidae, cerambycidae) were only found in MOCH diets, suggesting more dietary diversity or specialization at high elevations

## Model Diet and Reproductive Metrics (2019–2020 only)

```{r}
rich_2020 <- rich_2020 %>%
  filter(!grepl("Sum", SampleID))  # removes the summary row

# (You can do the same for rich_2019 if needed)


# Join diet richness to metadata using SampleID
model_data <- left_join(col_data, bind_rows(rich_2019, rich_2020), by = "SampleID")

# Join reproductive success data based on NestID (individuals → nest-level metadata)
model_data <- left_join(model_data, master_data, by = c("NestID" = "id"))

fitness_data_clean <- model_data %>%
  filter(!is.na(avg_tarsus), !is.na(diet_richness), !is.na(Species), !is.na(Elevation))

if (nrow(fitness_data_clean) > 0) {
  fitness_data_clean$Species <- as.factor(fitness_data_clean$Species)

  lm1 <- lm(avg_tarsus ~ diet_richness + Species + Elevation, data = fitness_data_clean)
  print(summary(lm1))

  lm2 <- lm(num_nestlings ~ diet_richness + Species + Elevation, data = fitness_data_clean)
  print(summary(lm2))
} else {
  message("No complete 2019–2020 data available for modeling.")
}
```

```{r}
## Model Diet and Reproductive Metrics (2024 only)

# Format 2024 NestIDs to match master_data (drop the middle digit)
model_data_2024 <- col_data %>%
  filter(Year == 2024) %>%
  mutate(NestID_clean = gsub("^(\\d+)-\\d+-2024$", "\\1-2024", NestID)) %>%
  left_join(r_df_2024, by = "SampleID") %>%
  left_join(master_data, by = c("NestID_clean" = "id"))

# Filter for complete data
fitness_data_2024 <- model_data_2024 %>%
  filter(!is.na(avg_tarsus), !is.na(diet_richness), !is.na(Species), !is.na(Elevation))

# Run models
if (nrow(fitness_data_2024) > 0) {
  fitness_data_2024$Species <- as.factor(fitness_data_2024$Species)

  lm1_2024 <- lm(avg_tarsus ~ diet_richness + Species + Elevation, data = fitness_data_2024)
  print(summary(lm1_2024))

  lm2_2024 <- lm(num_nestlings ~ diet_richness + Species + Elevation, data = fitness_data_2024)
  print(summary(lm2_2024))
} else {
  message("No complete 2024 data available for modeling.")
}


```


## Conservation Metric: Invasive Species (2024 only)

```{r}
invasive_list <- c("Harmonia axyridis", "Coccinella septempunctata")
invasive_ids <- taxonomy %>% filter(Species %in% invasive_list) %>% pull(ESVId)
esv_2024 <- esv_2024[, !is.na(names(esv_2024)) & names(esv_2024) != ""]


inv_counts <- esv_2024 %>%
  filter(ESVId %in% invasive_ids) %>%
  select(ESVId, all_of(sample_cols_2024)) %>%
  pivot_longer(cols = -ESVId, names_to = "SampleID", values_to = "count") %>%
  filter(count > 0) %>%
  group_by(SampleID) %>%
  summarise(invasive_count = n())

inv_data <- left_join(col_data, inv_counts, by = "SampleID")
inv_data$invasive_count[is.na(inv_data$invasive_count)] <- 0

ggplot(inv_data, aes(x = Species, y = invasive_count, fill = Species)) +
  geom_boxplot() +
  labs(title = "Invasive Prey in Diet by Species", y = "# Invasive Prey Items") +
  theme_minimal()
```

## Reflection Questions

- How do diets differ in diversity and taxonomic composition between BCCH and MOCH?
- Are chickadees consuming invasive species more in certain habitats?
- Does higher diet richness correlate with better reproductive outcomes?

## Suggested Readings

- Grabenstein et al. (2022). Urban-rural gradients and chickadee diet.
- Norris et al. (2022). Insect availability and fecundity in montane birds.
- Trevelline et al. (2018). Nestling provisioning and niche separation.

