---
title: "Module 3"
author: "Sara Padula"
date: "2025-05-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project Goals

- Compare diet composition between black-capped and mountain chickadees.
- Assess how diet varies across years (2019, 2020, 2024) and elevation.
- Visualize dietary overlap using trees and ordination.
- Model variation in diet richness by species, year, and elevation.

```{r}

## ---- Load libraries ----
library(dplyr)
library(tidyr)
library(vegan)
library(ape)
library(ggplot2)
library(ggtree)

# ---- Load ColData and all ESVs ----

col_data <- read.csv("ColData.csv", header = TRUE)
esv.2024 <- read.csv("JVB4821-ArthCOI-read-data.csv", header = TRUE, check.names = FALSE)
esv.2019.raw <- read.csv("ESV_Long_2019_Ind.csv", header = TRUE, check.names = FALSE)
esv.2020.raw <- read.csv("ESV_Long_2020_Ind.csv", header = TRUE, check.names = FALSE)

# ---- Helper function to process ESV data ----

process_esv <- function(raw_esv) {
  # Get all columns except the first (assumed to be ESV ID)
  sample_cols <- names(raw_esv)[-1]
  
  # Create valid column names from ESV IDs
  esv_ids <- make.names(raw_esv[[1]], unique = TRUE)
  
  # Transpose and label
  transposed <- setNames(data.frame(t(raw_esv[, sample_cols])), esv_ids)
  rownames(transposed) <- sample_cols
  
  # Clean values
  transposed <- transposed %>% mutate(across(everything(), ~as.numeric(.)))
  transposed[is.na(transposed)] <- 0
  
  # Filter empty samples and create binary matrix
  trans_clean <- transposed[rowSums(transposed) > 0, ]
  bin_matrix <- (trans_clean > 0) * 1
  
  return(list(trans = transposed, clean = trans_clean, bin = bin_matrix))
}



# ---- Process ESV data for each year ----

# 2019
esv.2019.proc <- process_esv(esv.2019.raw)
esv.2019.trans <- esv.2019.proc$trans
esv.2019.trans.clean <- esv.2019.proc$clean
esv.2019.bin <- esv.2019.proc$bin

# 2020
esv.2020.proc <- process_esv(esv.2020.raw)
esv.2020.trans <- esv.2020.proc$trans
esv.2020.trans.clean <- esv.2020.proc$clean
esv.2020.bin <- esv.2020.proc$bin

# 2024
sample_cols_2024 <- names(esv.2024)[grepl("^S\\d+", names(esv.2024))]
esv.2024.trans <- setNames(data.frame(t(esv.2024[, sample_cols_2024])), esv.2024$ESVId)
rownames(esv.2024.trans) <- gsub("^X", "", rownames(esv.2024.trans))
rownames(esv.2024.trans) <- trimws(rownames(esv.2024.trans))
esv.2024.trans <- esv.2024.trans %>% mutate(across(everything(), ~as.numeric(.)))
esv.2024.trans[is.na(esv.2024.trans)] <- 0
esv.2024.bin <- (esv.2024.trans > 0) * 1

# ---- Clean and match metadata ----

# Clean ColData SampleIDs
col_data$SampleID <- gsub("\\.1$", "", col_data$SampleID)
col_data$SampleID <- gsub("^S", "", col_data$SampleID)
col_data$SampleID <- gsub(" ", "", col_data$SampleID)

# 2024 metadata
sample_ids_clean <- gsub("\\.1$", "", gsub("^S", "", rownames(esv.2024.trans)))
meta.2024 <- col_data %>%
  filter(Year == 2024, SampleID %in% sample_ids_clean)
meta.2024$SampleID <- paste0("S", meta.2024$SampleID, ".1")
rownames(meta.2024) <- meta.2024$SampleID
stopifnot(all(rownames(meta.2024) %in% rownames(esv.2024.trans)))

# 2019 metadata
sample_ids_2019 <- rownames(esv.2019.trans)
meta.2019 <- col_data %>% filter(Year == 2019, SampleID %in% sample_ids_2019)
rownames(meta.2019) <- meta.2019$SampleID
stopifnot(all(rownames(meta.2019) %in% rownames(esv.2019.trans)))

# 2020 metadata
sample_ids_2020 <- rownames(esv.2020.trans)
meta.2020 <- col_data %>% filter(Year == 2020, SampleID %in% sample_ids_2020)
rownames(meta.2020) <- meta.2020$SampleID
stopifnot(all(rownames(meta.2020) %in% rownames(esv.2020.trans)))


```

## Tree-Based Clustering (Jaccard)

```{r jaccard-trees}
# Jaccard distances
jaccard.2019 <- dist(esv.2019.bin, method = "binary")
jaccard.2020 <- dist(esv.2020.bin, method = "binary")
jaccard.2024 <- dist(esv.2024.bin, method = "binary")

# Neighbor-joining trees
tree.2019 <- nj(jaccard.2019)
tree.2020 <- nj(jaccard.2020)
tree.2024 <- nj(jaccard.2024)

options(ignore.negative.edge = TRUE)

# Plots
p1 <- ggtree(tree.2019) %<+% meta.2019 +
  geom_tippoint(aes(color = Species)) +
  labs(title = "Jaccard Tree - 2019") +
  scale_color_manual(values = c("rosybrown2", "cornflowerblue"))

p2 <- ggtree(tree.2020) %<+% meta.2020 +
  geom_tippoint(aes(color = Species)) +
  labs(title = "Jaccard Tree - 2020") +
  scale_color_manual(values = c("rosybrown2", "cornflowerblue"))

# Ensure rownames in meta match those in esv.2024.bin
rownames(meta.2024) <- meta.2024$SampleID

# Rebuild Jaccard tree
jaccard.2024 <- dist(esv.2024.bin, method = "binary")
tree.2024 <- nj(jaccard.2024)

# Plot with metadata joined and color by Species
p3 <- ggtree(tree.2024) %<+% meta.2024 +
  geom_tippoint(aes(color = Species)) +
  labs(title = "Jaccard Tree - 2024") +
  scale_color_manual(values = c("rosybrown2", "cornflowerblue"))

print(p1); print(p2); print(p3)
```

Jaccard tree interpretation:
Each tip of the tree = one nest's diet... nests are grouped based on how similar the insects in their fecal samples were

2019
-fewer samples overall
-BCCH and MOCh diets are somewhat mixed but show some species-specific clustering
-a good baseline for comparison

2020
-more samples than 2019
-a lot of spread... nest diets varied a lot across individuals
a few samples have missing species metadata

2024
-stronger clustering by species than in 2020
-many MOCh samples group together, suggesting they were eating similar things
-BCChs appear mostly in a separate bunch, which could point to species-level differences in diet

## NMDS Ordination (Bray-Curtis)

```{r nmds-ordination}
# Clean 2020 of empty rows
esv.2020.trans.clean <- esv.2020.trans[rowSums(esv.2020.trans) > 0, ]
meta.2020.clean <- meta.2020 %>% filter(SampleID %in% rownames(esv.2020.trans.clean))

# Distances
bray.2019 <- vegdist(esv.2019.trans, method = "bray")
bray.2020 <- vegdist(esv.2020.trans.clean, method = "bray")
bray.2024 <- vegdist(esv.2024.trans, method = "bray")

# NMDS
nmds.2019 <- metaMDS(bray.2019, k = 2)
nmds.2020 <- metaMDS(bray.2020, k = 2)
nmds.2024 <- metaMDS(bray.2024, k = 2)

# Dataframes
nmds_df_2019 <- data.frame(nmds.2019$points, SampleID = rownames(nmds.2019$points)) %>%
  left_join(meta.2019, by = "SampleID")

nmds_df_2020 <- data.frame(nmds.2020$points, SampleID = rownames(nmds.2020$points)) %>%
  left_join(meta.2020.clean, by = "SampleID")

nmds_df_2024 <- data.frame(nmds.2024$points, SampleID = rownames(nmds.2024$points)) %>%
  left_join(meta.2024, by = "SampleID")

# Plots
ggplot(nmds_df_2019, aes(MDS1, MDS2, color = Species)) +
  geom_point(size = 3) + theme_classic() +
  labs(title = "NMDS - 2019") +
  scale_color_manual(values = c("rosybrown2", "cornflowerblue"))

ggplot(nmds_df_2020, aes(MDS1, MDS2, color = Species)) +
  geom_point(size = 3) + theme_classic() +
  labs(title = "NMDS - 2020") +
  scale_color_manual(values = c("rosybrown2", "cornflowerblue"))

ggplot(nmds_df_2024, aes(MDS1, MDS2, color = Species)) +
  geom_point(size = 3) + theme_classic() +
  labs(title = "NMDS - 2024") +
  scale_color_manual(values = c("rosybrown2", "cornflowerblue"))
```

What the NMDS Plots Show

Each dot = a nest. Nests close together had more similar diets (based on insect DNA in the feces). Dots further apart = more different diets.

2019
-patterns are hard to see
-large distances between dots might suggest high variability, or just low sample size

2020
-no strong clustering by species
-both MOCH and BCCH are spread out and overlap a lot
-diets in 2020 were highly variable, and species may have been eating similar things
-possibly higher food diversity or less competition in this year

2024
-clear separation between MOCH and BCCH diets
-MOCH diets are tightly clustered --> many individuals eat similar prey
-BCCH diets are more spread out --> more diet variety
-Suggests species may be specializng on different insect groups

## Load Libraries

```{r}
library(phyloseq)
library(vegan)
library(tidyr)
library(dplyr)
library(ggplot2)
```

### ------- Alpha Diversity -------
```{r}
library(phyloseq)

### ----- 2019 ----- ###
# Set rownames for metadata
rownames(meta.2019) <- meta.2019$SampleID

# Keep only metadata rows that match the ESV table
meta.2019.filtered <- meta.2019[rownames(esv.2019.bin), , drop = FALSE]

# Construct phyloseq object
otu_2019 <- otu_table(as.matrix(esv.2019.bin), taxa_are_rows = FALSE)
samp_2019 <- sample_data(meta.2019.filtered)
ps_2019 <- phyloseq(otu_2019, samp_2019)


### ----- 2020 ----- ###
# Set rownames for metadata
rownames(meta.2020.clean) <- meta.2020.clean$SampleID

# Keep only metadata rows that match the cleaned ESV table
meta.2020.filtered <- meta.2020.clean[rownames(esv.2020.trans.clean), , drop = FALSE]

# Construct phyloseq object
otu_2020 <- otu_table(as.matrix(esv.2020.trans.clean), taxa_are_rows = FALSE)
samp_2020 <- sample_data(meta.2020.filtered)
ps_2020 <- phyloseq(otu_2020, samp_2020)


### ----- 2024 ----- ###
# Get all sample IDs from ESV matrix
sample_ids_2024 <- rownames(esv.2024.bin)  # e.g., "S115308.1"

# Remove trailing ".1" and leading "S" to get base IDs
base_ids_2024 <- gsub("\\.1$", "", gsub("^S", "", sample_ids_2024))

# Filter metadata to 2024 and match base IDs
meta.2024 <- col_data %>%
  filter(Year == 2024, SampleID %in% base_ids_2024)

# Now reformat SampleID to match rownames in ESV matrix
meta.2024$SampleID <- paste0("S", meta.2024$SampleID, ".1")
rownames(meta.2024) <- meta.2024$SampleID

# Final match filter (just in case)
meta.2024.filtered <- meta.2024[rownames(meta.2024) %in% sample_ids_2024, ]

# Build phyloseq object
otu_2024 <- otu_table(as.matrix(esv.2024.bin), taxa_are_rows = FALSE)
samp_2024 <- sample_data(meta.2024.filtered)
ps_2024 <- phyloseq(otu_2024, samp_2024)


# Richness plots
plot_richness(ps_2019, x = "Species", measures = c("Observed", "Shannon")) +
  geom_boxplot(alpha = 0.3) + theme_classic() + labs(title = "Alpha Diversity - 2019")

plot_richness(ps_2020, x = "Species", measures = c("Observed", "Shannon")) +
  geom_boxplot(alpha = 0.3) + theme_classic() + labs(title = "Alpha Diversity - 2020")

plot_richness(ps_2024, x = "Species", measures = c("Observed", "Shannon")) +
  geom_boxplot(alpha = 0.3) + theme_classic() + labs(title = "Alpha Diversity - 2024")
```

Alpha Diversity interpretation

2019
-Observed richness: BCCHs had a wider range of prey types compared to MOCHs.
-Shannon diversity: BCCHs also showed greater variation in evenness -- some had very diverse and balanced diets, while others were less so
-Takeaway: In 2019, BCCH diets were slightly more diverse and variable overall than MOCH diets

2020
-Observed richness: MOCHs trended slightly higher in richness, but both species showed relatively low diversity
-Shannon diversity: Diets were overall less even (lower shannon values), especially for BCCHs-- some fed mainly on just a few prey types
-Takeaway: 2020 appears to be a year of low diversity overall - could suggest food scarcity or strong prey preference

2024
Observed richness: MOCHs had much higher and more variable richness -- some nests had >90 prey types
Shannon diversity: MOCHs again stood out -- higher evenness and overall dietary diversity compared to BCCHs
Takeaway: In 2024, MOCH diets were richer and more evenly spread across prey types. BCCHs were more specialized or limited.

### ------- PERMANOVA ------
```{r}
## PERMANOVA (adonis2) for Species Effects on Diet Composition (Bray-Curtis)

# ---- 2019 ----
# Subset metadata to match distance matrix
meta.2019.sub <- meta.2019[match(rownames(esv.2019.trans), meta.2019$SampleID), ]
meta.2019.sub <- meta.2019.sub[!is.na(meta.2019.sub$Species), ]

# Subset Bray-Curtis matrix
bray.2019.mat <- as.matrix(bray.2019)
bray.2019.filtered <- as.dist(bray.2019.mat[rownames(meta.2019.sub), rownames(meta.2019.sub)])

# Run PERMANOVA
adonis_2019 <- adonis2(bray.2019.filtered ~ Species, data = meta.2019.sub)
print(adonis_2019)


# ---- 2020 ----
# Subset metadata for cleaned 2020 data
meta.2020.sub <- meta.2020.clean[match(rownames(esv.2020.trans.clean), meta.2020.clean$SampleID), ]
meta.2020.sub <- meta.2020.sub[!is.na(meta.2020.sub$Species), ]

# Subset Bray-Curtis matrix
bray.2020.mat <- as.matrix(bray.2020)
bray.2020.filtered <- as.dist(bray.2020.mat[rownames(meta.2020.sub), rownames(meta.2020.sub)])

# Run PERMANOVA
adonis_2020 <- adonis2(bray.2020.filtered ~ Species, data = meta.2020.sub)
print(adonis_2020)

# ---- 2024 ----
# Use already cleaned metadata and ESV matrix
# meta.2024.filtered and esv.2024.trans should be ready

# Double-check rownames match
valid_samples_2024 <- intersect(rownames(esv.2024.trans), rownames(meta.2024.filtered))

# Filter to matched rows
esv.2024.filtered <- esv.2024.trans[valid_samples_2024, ]
meta.2024.filtered <- meta.2024.filtered[valid_samples_2024, ]

# Drop NAs and ensure ≥2 individuals per species
meta.2024.sub <- meta.2024.filtered[!is.na(meta.2024.filtered$Species), ]
meta.2024.sub <- meta.2024.sub[meta.2024.sub$Species %in% names(table(meta.2024.sub$Species))[table(meta.2024.sub$Species) > 1], ]

# Filter ESV matrix accordingly
esv.2024.sub <- esv.2024.filtered[rownames(meta.2024.sub), ]

# Run Bray-Curtis
bray.2024.filtered <- vegdist(esv.2024.sub, method = "bray")

# Run PERMANOVA
adonis_2024 <- adonis2(bray.2024.filtered ~ Species, data = meta.2024.sub)

print(adonis_2019)
print(adonis_2020)
print(adonis_2024)
```
What does this test do?
Does diet composition differ between black-capped chickadees and mountain chickadees? It answers this question by using Bray-Curtis distances to compare how different individuals' diets are from each other.

2019 --

p = 0.06 --> almost statistically signficiant but not quite
-chickadees might have slightly different diets but with weak evidence

2020 --
- p = 0.011 --> statistically significant
- chickadee species had meaningfully different diets
-R^2 = 0.038 --> species explains ~4% of the variation in diet

2024 -- 
- p = 0.001 --> highly significant difference in diet by species
- R^2 = 0.052 --> species explains 5.2% of the variation in diet

Across all three years, there’s increasing evidence that black-capped and mountain chickadees differ in what they feed their babies, with the strongest difference in 2024.

### ------- Beta Dispersion -------
```{r}
## --- Beta Dispersion and PERMDISP Tests by Species --- ##

# --- 2019 ---
samples_2019 <- rownames(as.matrix(bray.2019))
meta.2019.bd <- meta.2019 %>% filter(SampleID %in% samples_2019)
meta.2019.bd <- meta.2019.bd[match(samples_2019, meta.2019.bd$SampleID), ]
bd_2019 <- betadisper(bray.2019, meta.2019.bd$Species)
anova(bd_2019)

# --- 2020 ---
samples_2020 <- rownames(as.matrix(bray.2020))
meta.2020.bd <- meta.2020.clean %>% filter(SampleID %in% samples_2020)
meta.2020.bd <- meta.2020.bd[match(samples_2020, meta.2020.bd$SampleID), ]
bd_2020 <- betadisper(bray.2020, meta.2020.bd$Species)
anova(bd_2020)

# 2024


meta.2024 <- col_data %>% filter(Year == 2024)

#  Format SampleIDs to match ESV table (add leading "S" if missing)
meta.2024$SampleID <- ifelse(grepl("^S", meta.2024$SampleID),
                             meta.2024$SampleID,
                             paste0("S", meta.2024$SampleID))

#  Append ".1" to match esv.2024.bin/bray.2024 rownames
meta.2024$SampleID <- paste0(meta.2024$SampleID, ".1")
rownames(meta.2024) <- meta.2024$SampleID

# 4. Remove NAs in Species
meta.2024.bd <- meta.2024[!is.na(meta.2024$Species), ]

#  Filter out species with only 1 individual
valid_species <- names(which(table(meta.2024.bd$Species) > 1))
meta.2024.bd <- meta.2024.bd[meta.2024.bd$Species %in% valid_species, ]

# Intersect with Bray-Curtis distance matrix
shared_ids <- intersect(rownames(as.matrix(bray.2024)), rownames(meta.2024.bd))
meta.2024.bd <- meta.2024.bd[shared_ids, ]

# Subset Bray-Curtis matrix
bray.2024.sub <- as.matrix(bray.2024)[shared_ids, shared_ids]
bray.2024.dist <- as.dist(bray.2024.sub)

# ---- Run betadisper ----
bd_2024 <- betadisper(bray.2024.dist, meta.2024.bd$Species)
anova(bd_2024)


```
This analysis is asking: Do black-capped and mountain chickadees differ in how variable their diets are from individual to individual within each species?

2019
-there is marginal diff in dispersion btwn species
-one species might have had slightly more variable diets but it isn't significant

2020
-there is no difference at all in the variability between species

2024
-no evidence for species differences in beta dispersion

### ------- Diet Composition by Species only 2024-------
```{r}
esv.2024.with.meta <- esv.2024.trans
esv.2024.with.meta$Species <- meta.2024$Species[match(rownames(esv.2024.trans), meta.2024$SampleID)]

mean_by_species_2024 <- esv.2024.with.meta %>%
  group_by(Species) %>%
  summarise(across(where(is.numeric), mean)) %>%
  pivot_longer(-Species, names_to = "ESV", values_to = "Mean_Abundance") %>%
  filter(Mean_Abundance > 0.01)  # adjust threshold if needed

head(mean_by_species_2024)

write.csv(mean_by_species_2024, "mean_by_species_2024.csv", row.names = FALSE)

```
### ------- Richness by Elevation Models -------
```{r}
# Calculate richness per sample
# ---- Match metadata rows to ESVs before calculating richness ----

# 2019
meta.2019 <- meta.2019[rownames(esv.2019.bin), , drop = FALSE]
meta.2019$Richness <- rowSums(esv.2019.bin)

# 2020
meta.2020.clean <- meta.2020.clean[rownames(esv.2020.trans.clean), , drop = FALSE]
meta.2020.clean$Richness <- rowSums(esv.2020.bin[rownames(esv.2020.trans.clean), ])

# 2024
meta.2024 <- meta.2024[rownames(esv.2024.bin), , drop = FALSE]
meta.2024$Richness <- rowSums(esv.2024.bin)

# ---- Models ----

mod_2019 <- lm(Richness ~ Elevation + Species, data = meta.2019)
mod_2020 <- lm(Richness ~ Elevation + Species, data = meta.2020.clean)
mod_2024 <- lm(Richness ~ Elevation + Species, data = meta.2024)

summary(mod_2019)
summary(mod_2020)
summary(mod_2024)

```
2019
-no significant effects of elevation or species

2020
-no significant effects but a marginal trend... higher elevation was associated with slightly greated richness (p=0.1, not significant)

2024
-neither variable was significant
## Reflection Questions

Species Differences in Diet
Across the three years, which year showed the clearest difference in diet between mountain and black-capped chickadees?

What might explain why the differences were stronger that year?

In 2024, MOCHs had diets that were more clustered together, while BCCHs were more spread out.
What might this suggest about their foraging behavior or specialization?

In the alpha diversity results, MOCHs had higher richness and evenness in 2024.
Why do you think MOCHs might be eating more different kinds of prey than BCCHs?


Effects of Elevation and Year
There was a weak trend that higher elevation was associated with more prey types in some years.

What ecological reasons might explain this?

Do you think year-to-year variation (like 2020 vs 2024) is mostly due to environmental changes (like weather, prey availability), or something about the birds themselves? Why?

Methodology and Interpretation
The PERMANOVA showed a significant difference in species diets in 2024, but low R² values (only ~5%). 

How would you explain to someone what it means when a result is significant but explains only a small amount of the variation?

The beta dispersion test looked at how variable diets were within species, but found no significant differences.

What might be the value of doing this test anyway?

Connecting to Your Own Project
Based on these analyses, are you more interested in:

How different species feed (species-level comparison)?

How diets change by elevation?

How diets vary across years?

Or how individuals within a species vary?

Is there another variable (like nestling mass, lay date, or habitat type) you’d want to explore in relation to diet? What would you predict?

What new hypothesis could you test using these same methods (trees, NMDS, richness models, etc.)?
